---
title: "Density, Distribution, Quantiles"
description: "Practical route to unconditional and conditional density/distribution workflows, plus nonparametric quantile regression in np."
keywords:
  - npudens
  - npudist
  - npcdens
  - npcdist
  - npqreg
  - density
  - quantile regression
---

```{r density-distribution-setup}
#| include: false
source("site_helpers.R", local = knitr::knit_global())
```

The `np` package is useful not only for regression but also for unconditional and conditional density estimation, distribution estimation, and nonparametric quantile regression. This page collects those workflows in one place.

If you want a minimal downloadable script for the simplest density workflow, start with [np_density_quickstart.R](www/np_npRmpi/np_density_quickstart.R).

If you want the shortest route to conditional density or quantile workflows, start with:

- [np_distribution_quickstart.R](www/np_npRmpi/np_distribution_quickstart.R)
- [np_conditional_density_quickstart.R](www/np_npRmpi/np_conditional_density_quickstart.R)
- [np_conditional_distribution_quickstart.R](www/np_npRmpi/np_conditional_distribution_quickstart.R)
- [np_quantile_quickstart.R](www/np_npRmpi/np_quantile_quickstart.R)

## Unconditional density and distribution: `npudens` and `npudist`

Use these when the object of interest is the joint or marginal distribution itself rather than a regression function.

The Old Faithful data remain a simple and informative illustration because the shape is not well summarized by a simple parametric family.

```r
## Fit unconditional density and distribution objects for the same data
library(np)
data(faithful, package = "datasets")

f_faithful <- npudens(~ eruptions + waiting, data = faithful)
F_faithful <- npudist(~ eruptions + waiting, data = faithful)

summary(f_faithful)
summary(F_faithful)
```

If you want to visualize the estimated surface, you can then plot the objects.

```r
## Plot the fitted unconditional objects to inspect their shape
plot(f_faithful, view = "fixed", main = "")
plot(F_faithful, view = "fixed", main = "")
```

This is a good example of why nonparametric methods can matter: a bimodal or otherwise irregular structure is much easier to reveal when you do not force a simple parametric family on the data.

## Conditional density and distribution: `npcdens` and `npcdist`

Use these when the full conditional distribution matters, not just the conditional mean.

The old GDP panel example remains helpful because the conditional distribution changes shape over time.

```r
## Fit conditional density and distribution objects given the covariate
library(np)
data(Italy, package = "np")

fhat <- npcdens(gdp ~ year, data = Italy)
Fhat <- npcdist(gdp ~ year, data = Italy)

summary(fhat)
summary(Fhat)
```

Plots can then be used to inspect the evolving conditional density and conditional distribution.

```r
## Plot the evolving conditional density and distribution surfaces
plot(fhat, view = "fixed", main = "")
plot(Fhat, view = "fixed", main = "")
```

This route is especially useful when the question is not simply whether the mean changes, but whether the entire distribution changes.

## Practitioner-oriented extras

Some applied tasks are a little different from the standard "fit and inspect"
workflow. The Cambridge practitioner material suggests three especially useful
ones:

- estimate a smooth probability function on discrete support,
- draw a smooth resample from a fitted unconditional density,
- draw a smooth resample from a fitted conditional density.

The first of these is compact enough to show inline.

Source file: [np_probability_quickstart.R](www/np_npRmpi/np_probability_quickstart.R)

```{r probability-quickstart-source}
#| echo: false
#| results: asis
emit_source("www/np_npRmpi/np_probability_quickstart.R")
```

Additional practitioner scripts:

- [np_density_resample_quickstart.R](www/np_npRmpi/np_density_resample_quickstart.R)
- [np_conditional_density_resample_quickstart.R](www/np_npRmpi/np_conditional_density_resample_quickstart.R)

## Copulas: `npcopula`

If the main object of interest is the dependence structure itself rather than a
regression function, `npcopula` is the natural route.

The formula interface is now the streamlined first stop: when `bws` is not
supplied, `npcopula()` automatically builds the required `npudistbw()`
bandwidth object for the default copula target, or `npudensbw()` when
`target = "density"`. For the default two-dimensional grid workflow, omitting
`u` also triggers automatic construction of a probability grid.

Source file: [np_copula_quickstart.R](www/np_npRmpi/np_copula_quickstart.R)

```{r copula-quickstart-source}
#| echo: false
#| results: asis
emit_source("www/np_npRmpi/np_copula_quickstart.R")
```

For LP-capable conditional families, the same modern shortcut is now available:

```r
## Use the same NOMAD shortcut for LP conditional density and distribution fits
fhat_lp <- npcdens(eruptions ~ waiting, data = dat, nomad = TRUE, nmulti = 1)
Fhat_lp <- npcdist(eruptions ~ waiting, data = dat, nomad = TRUE, nmulti = 1)
```

That is the convenience route for automatic local-polynomial degree and
bandwidth search in a single call. As with regression, it requires the `crs`
package because NOMAD degree search is provided there.

When the local-polynomial degree is above zero, `proper = TRUE` is also
available in the `npc*` calls and related plot routes. This enforces proper
conditional density/distribution output in the practical sense used here:
non-negative and integrating to one conditional on `x`.

## Nonparametric quantile regression: `npqreg`

Use `npqreg` when your interest lies in conditional quantiles rather than means.

One natural workflow is:

1. compute a conditional-distribution bandwidth object,
2. extract one or more conditional quantiles from that object.

The formula interface below is the streamlined version of that workflow:
when `bws` is not supplied, `npqreg()` automatically calls `npcdistbw()` to
select the conditional-distribution bandwidth object first. Bandwidth-selection
options supplied to `npqreg()` are passed through to that internal `npcdistbw()`
call, so you can still control the underlying conditional-distribution fit.

```r
## Extract several quantiles from one selected conditional-distribution fit
library(np)
data(Italy, package = "np")

qfit <- npqreg(gdp ~ ordered(year),
               data = Italy,
               tau = c(0.25, 0.50, 0.75))
```

Then compare the fitted quantiles:

```r
## Overlay the fitted conditional quantiles
plot(qfit)
```

If you first computed only one quantile, reuse the selected bandwidths for
additional quantiles without recomputing cross-validation. Call `npqreg()` when
you need the fitted values as an object, or pass `tau` directly to `plot()` when
graphical inspection is all you need:

```r
qmed <- npqreg(gdp ~ ordered(year), data = Italy, tau = 0.50)
qfit <- npqreg(bws = qmed$bws, tau = c(0.25, 0.50, 0.75))
plot(qfit)

## Or, if plotting is the only goal:
plot(qmed, tau = c(0.25, 0.50, 0.75))
```

If you want derivative information, request it in the fit:

```r
qgrad <- npqreg(gdp ~ ordered(year),
                data = Italy,
                tau = 0.50,
                gradients = TRUE)

plot(qgrad, gradients = TRUE)
```

The same formula interface can also use the automatic local-polynomial
degree-and-bandwidth search when the conditioning data include at least one
continuous variable:

```r
Italy$year_numeric <- as.numeric(as.character(Italy$year))

qfit_lp <- npqreg(gdp ~ year_numeric,
                  data = Italy,
                  tau = c(0.25, 0.50, 0.75),
                  nomad = TRUE)
```

This is a good example of when it pays to separate bandwidth selection from later estimation rather than recomputing the same object repeatedly.

## When should I use these instead of regression?

| If you want... | Start with |
|---|---|
| A regression function or conditional mean | `npreg` |
| An unconditional density or distribution | `npudens`, `npudist` |
| A conditional density or conditional distribution | `npcdens`, `npcdist` |
| Conditional quantiles | `npqreg` |

## Practical advice

- density and conditional-density workflows can be computationally heavier than users expect,
- use explicit bandwidth objects when reusing the same structure across multiple fits,
- plots are often the most informative part of the workflow,
- if the run becomes expensive and the workflow is stable, that may be a reason to move to `npRmpi`.

## What to read next

- [Kernel Primer](kernel_primer.qmd)
- [Entropy and Testing](entropy_tests.qmd)
- [Semiparametric Models](semiparametric_models.qmd)
- [MPI and Large Data](mpi_large_data.qmd)
