---
title: "Interactive Demos"
description: "RStudio-first interactive demos for density, regression, and constrained estimation with np and crs."
keywords:
  - interactive demos
  - manipulate
  - RStudio
  - np
  - crs
---

```{r interactive-setup}
#| include: false
source("site_helpers.R", local = knitr::knit_global())
```

These examples are primarily pedagogical. Most use the `manipulate` package and are intended to be run inside RStudio so that you can move sliders, change kernels, alter smoothing parameters, and immediately see the effect on the estimate.

## A quick way to run one

Download or open one of the scripts below in RStudio, then source it. Once the plot appears, click the small tool icon in the plot pane if the control panel is not already visible.

## Best first demos

| If you want to see... | Start here | Notes |
|---|---|---|
| How bandwidth, kernel, and sample size change a density estimate | [manipulate_density.R](www/Fun/manipulate_density.R) | good first teaching demo |
| How off-axis quantiles change multivariate partial plots | [manipulate_wage1.R](www/Fun/manipulate_wage1.R) | uses the `wage1` data and precomputed bandwidth object |
| Shape restrictions in action | [manipulate_constrained_spline.R](www/Fun/manipulate_constrained_spline.R) | requires `quadprog` |

## Density and distribution demos

- Univariate density with bandwidth, kernel, and sample-size controls: [manipulate_density.R](www/Fun/manipulate_density.R)
- Old Faithful univariate density: [manipulate_eruptions.R](www/Fun/manipulate_eruptions.R)
- Univariate distribution function: [manipulate_distribution.R](www/Fun/manipulate_distribution.R)
- Bivariate density: [manipulate_bivariate_density.R](www/Fun/manipulate_bivariate_density.R)
- Old Faithful bivariate density: [manipulate_faithful_density.R](www/Fun/manipulate_faithful_density.R)
- Bivariate distribution function: [manipulate_bivariate_distribution.R](www/Fun/manipulate_bivariate_distribution.R)

## Regression and shape-restriction demos

- Multivariate regression partial plots: [manipulate_wage1.R](www/Fun/manipulate_wage1.R)
- Constrained local polynomial regression: [manipulate_constrained_local_polynomial.R](www/Fun/manipulate_constrained_local_polynomial.R)
- Constrained production function example: [manipulate_constrained_local_polynomial_production.R](www/Fun/manipulate_constrained_local_polynomial_production.R)
- Constrained spline illustrations: [manipulate_constrained_spline.R](www/Fun/manipulate_constrained_spline.R), [manipulate_constrained_spline_derivative.R](www/Fun/manipulate_constrained_spline_derivative.R)

## Other demos

- Copula plots: [manipulate_copula.R](www/Fun/manipulate_copula.R)
- SOCCO example: [manipulate_socco.R](www/Fun/manipulate_socco.R), [data.RData](www/Fun/data.RData), [figures.R](www/Fun/figures.R)

## Notes

- Some scripts require additional packages such as `quadprog`, `mnormt`, or `rgl`.
- The older `npglpreg` route is deprecated. For a functional replacement, use `npreg(..., nomad = TRUE)` in `np` or `npRmpi`.
- The `manipulate` package is tied to RStudio, so these are best treated as RStudio-first examples rather than generic batch scripts.

## Inline source for two flagship demos

::: {.callout-note collapse="true"}
## Density demo source

Source file: [manipulate_density.R](www/Fun/manipulate_density.R)

```{r density-demo-source}
#| echo: false
#| results: asis
emit_source("www/Fun/manipulate_density.R")
```
:::

::: {.callout-note collapse="true"}
## Multivariate wage demo source

Source file: [manipulate_wage1.R](www/Fun/manipulate_wage1.R)

```{r wage-demo-source}
#| echo: false
#| results: asis
emit_source("www/Fun/manipulate_wage1.R")
```
:::
