---
title: "Quickstarts"
description: "Small runnable scripts for the most common np, npRmpi, and crs entry points."
keywords:
  - quickstart
  - np
  - npRmpi
  - crs
  - examples
  - starter scripts
---

```{r quickstarts-setup}
#| include: false
source("site_helpers.R", local = knitr::knit_global())
quickstarts <- read_quickstarts_manifest()
```

This page is the shortest route from “I need an example” to a small runnable script. The scripts below are the modern quickstarts in the gallery, and each block is pulled directly from the underlying source file.

## Choose a starting point

```{r quickstart-chooser}
#| echo: false
#| results: asis
render_quickstart_choose_table(quickstarts)
```

## `np` quickstarts

```{r np-quickstarts-callouts}
#| echo: false
#| results: asis
render_quickstart_callouts(subset(quickstarts, section == "np"))
```

## `npRmpi` quickstarts

```{r nprmpi-quickstarts-callouts}
#| echo: false
#| results: asis
render_quickstart_callouts(subset(quickstarts, section == "npRmpi"))
```

## `crs` quickstart

```{r crs-quickstarts-callouts}
#| echo: false
#| results: asis
render_quickstart_callouts(subset(quickstarts, section == "crs"))
```

## Where to go next

- [Code Catalog](code_catalog.qmd) for the broader script library
- [Kernel Methods](np_npRmpi.qmd) for `np`
- [Density, Distribution, Quantiles](density_distribution_quantiles.qmd) for conditional densities, distributions, and quantiles
- [MPI and Large Data](mpi_large_data.qmd) for `npRmpi`
- [Splines](crs.qmd) for `crs`
